\bf Model Visualization Techniques for a Social Network Model

### 2017-03-30 paper.Rnw
some adj mat stuff

A key component of each step of the SIENA method of moments algorithm is the ``microstep." This step simulates from the model in its current state, $x(t_m)$ with current parameter values $\theta_{0}$, to the next state, $x(t_{m+1})$. This microstep process stops when the simulated network has acheived the same number of differences from $x(t_m)$ as $x(t_{m+1})$. This simulation process follows the steps of the continuous-time Markov chain. First, an ``ego node'' is selected to make a change, and then it randomly makes one change in its ties according to the probability, $p_{ij}$, determined by its objective function. The options for change are (1) removing a current tie, (2) adding a new tie, or (3) making no change at all. Saving all of these steps is not efficient if you're only interested in parameter values, but they can be saved and extracted using options in \texttt{RSiena}, which is exactly what we did to create our visualizations. Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps. We wanted to see these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain.

@@

>>>>

A key component of each step of the SIENA method of moments algorithm is the ``microstep." This step simulates from the model in its current state, $x(t_m)$ with current parameter values $\theta_{0}$, to the next state, $x(t_{m+1})$. This microstep process stops when the simulated network has acheived the same number of differences, $C$, from $x(t_m)$ as $x(t_{m+1})$, where 
$$C = \sum_{i \neq j} |x_{ij}(t_{m+1}) - x_{ij}(t_{m})|.$$ 
This simulation process follows the steps of the continuous-time Markov chain. First, an ``ego node'' is selected to make a change, and then it randomly makes one change in its ties according to the probability, $p_{ij}$, determined by its objective function. The options for change are (1) removing a current tie, (2) adding a new tie, or (3) making no change at all. Saving all of these steps is not efficient if you're only interested in parameter values, but they can be saved and extracted using options in \texttt{RSiena}, which is exactly what we did to create our visualizations. Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps. We wanted to see these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain.

@@

XXX HH's code XXX

INSERT DESCRIPTION OF THE MOVIE - WHAT IT ALLOWS US TO SEE, ETC.

We also use animation to view the changing structure of the adjacency matrix the microsteps. The adjacency matrices for the three waves of friendship data as shown in Figure~\ref{fig:alldataviz} are ordered by node id. There are 50 nodes in the data, numbered 1-50, and that order is used on the $x$ and $y$ axes for the matrix visualization. Viewing the adjacency matrices with this arbitrary ordering does not provide much information to the viewer about the underlying structure of the network. This lack of perceived structure would be exacerbated in an animation. So, we adjust the ordering so that the viewer can perceive more of the structure of the graph. This process is known as matrix seriation \citep{seriation}.

To reorder the vertices for the matrix visualization, we first constructed a cumulative adjaceny matrix, $\mathbf{A}^{cum}$, for the series of microsteps simulating the network from $x(t_m)$ to $x(t_{m+1})$. A single entry in the cumulative adjacency matrix, $\mathbf{A}^{cum}_{ij}$, is the total number of times the edge from node $i$ to node $j$ appears in the network from the initial observation, $x(t_m) \equiv X(0)$ to the final reult of the last microstep, $X(C)$: 
$$ \mathbf{A}^{cum}_{ij} = \sum_{c = 0}^C X_{ij}(c). $$
We then performed a principal component analysis on $\mathbf{A}^{cum}$, and used the values of the first principal component to order the vertices on the $x$ and $y$ axes for the adjacency matrix animation.
